---
title: "Midterm "
author: "hp621@scarletmail.rutgers.edu"
date: "04/01/2023"
output: html_document
---
```{r}
library(factoextra)
library(corrplot)
library(cluster)
library(psych)
library(dplyr)
library(devtools)
library(MASS)
library(Hotelling)
library(ggplot2)
library(tidyr)
library(readr)
library(FactoMineR)
library(GGally)
library("factoextra")

```
## Importing Data and initial analysis
## Importing csv file

```{r}
Animal_Rights <- read_csv("C:/Users/HARSH/Desktop/RU/SEM 2/MVA/Assignments/Animal_Rights.csv")
View(Animal_Rights)
```


•	Describe the data and its values briefly. 
```{r}
dim(Animal_Rights)
#It has 67 rows and 11 columns

colnames(Animal_Rights)
#The name of following columns of the dataset


str(Animal_Rights)

#It shows the columns country is in character while others are in numeric data


summary(Animal_Rights)

attach(Animal_Rights)

```

#On average, 71.64% of countries recognize animal sentience, while 89.55% acknowledge animal suffering.
#Around 97.01% of countries have legislation against animal cruelty, indicating a widespread legal framework for animal protection.
#The majority of countries (97.01%) support the Universal Declaration of Animal Welfare, with 89.55% implementing a fur-farming ban.
#Median meat consumption per capita is 65.77 kilograms, highlighting significant consumption levels.
#Despite efforts with protected areas (median: 19.69% of land), pesticide use remains high (mean: 3.199 kilograms), suggesting potential environmental concerns.


```{r}
#Visualize the countries across all the dimensions in one chart so you can easily see similar countries. 

ggscatmat(Animal_Rights, columns=2:11, color="Country")

stars(Animal_Rights, columns=2:11,labels=Animal_Rights$Country)
```

```{r}
#Compute the distance between the countries across the different dimensions.

scale <- scale(Animal_Rights[-1])
#mean of each column
animalmean <- colMeans(Animal_Rights[-1])
#correlation of each column
animalcor <- cor(Animal_Rights[-1])
#covariance
animalcov <- cov(Animal_Rights[-1])



# Computing the Mahalanobis distance for each country
mahalanobis_distance <- function(x) {
  mahalanobis(x, animalmean, animalcov)
}

# Applying the Mahalanobis distance function to each row of the dataset
distance_matrix <- apply(Animal_Rights[-1], 1, mahalanobis_distance)

# Converting the distance matrix to a data frame for better readability
distance_df <- data.frame(Country , Distance = distance_matrix)

# Displaying the Mahalanobis distance for each country
print(distance_df)

```
#These are the distance of each country across the dimensions


```{r}
#Carry out cluster analysis to study the relationship between countries. Explain the number of clusters you chose as optimal and why. Show visualizations to support your answers where possible. 



scale <- scale(Animal_Rights[-1])

rownames(scale) <- Animal_Rights$Country

# we take 2 clusters, k-means = 2
kmeans2.animals <- kmeans(scale,2,nstart = 10)

# Computing the percentage of variation accounted for two  clusters
perc.var.2 <- round(100*(1 - kmeans2.animals$betweenss/kmeans2.animals$totss),1)
names(perc.var.2) <- "Perc. 2 clus"
perc.var.2

# we take 3 clusters, k-means = 3
kmeans3.animals <- kmeans(scale,3,nstart = 10)
# Computing the percentage of variation accounted for three  clusters
perc.var.3 <- round(100*(1 - kmeans3.animals$betweenss/kmeans3.animals$totss),1)
names(perc.var.3) <- "Perc. 3 clus"
perc.var.3

# For 4 clusters, k-means = 4
kmeans4.animals <- kmeans(scale,4,nstart = 10)
# Computing the percentage of variation accounted for four clusters
perc.var.4 <- round(100*(1 - kmeans4.animals$betweenss/kmeans4.animals$totss),1)
names(perc.var.4) <- "Perc. 4 clus"
perc.var.4

#For 5 clusters, k-means = 5
kmeans5.animals <- kmeans(scale,5,nstart = 10)
# Computing the percentage of variation accounted for four clusters
perc.var.5 <- round(100*(1 - kmeans5.animals$betweenss/kmeans5.animals$totss),1)
names(perc.var.5) <- "Perc. 5 clus"
perc.var.5

```
#The 2 cluster shows the highest variance so we go with 2 clusters
```{r}
# We divide the dataset into three clusters.
# Filtering properties which are in 1 cluster of k mean 3
clus1 <- matrix(names(kmeans3.animals$cluster[kmeans3.animals$cluster == 1]),
                ncol=1, nrow=length(kmeans3.animals$cluster[kmeans3.animals$cluster == 1]))
colnames(clus1) <- "Cluster 1"
clus1

# Filtering properties which are in 2 cluster of k mean 3
clus2 <- matrix(names(kmeans2.animals$cluster[kmeans2.animals$cluster == 2]),
                ncol=1, nrow=length(kmeans2.animals$cluster[kmeans2.animals$cluster == 2]))
colnames(clus2) <- "Cluster 2"
clus2

# Filtering properties which are in 3 cluster of k mean 3
clus3 <- matrix(names(kmeans3.animals$cluster[kmeans3.animals$cluster == 3]),
                ncol=1, nrow=length(kmeans3.animals$cluster[kmeans2.animals$cluster == 3]))
colnames(clus3) <- "Cluster 3"
clus3
```
```{r}
# We can use the hierarchical clustering method, which allows us to create a dendrogram 
# to visualize the relationships between the countries based on their animal rights
# Hierarchical Clustering

# Calculating the distance matrix using Euclidean distance
dist.animals <- dist(Animal_Rights, method = "euclidean") # Distance matrix

# Performing hierarchical clustering using complete linkage
fit <- hclust(dist.animals, method="complete")

#Plotting the dendogram

plot(fit, main="Dendrogram of European countries based on diet")

#Cutting the tree into 3 clusters
groups <- cutree(fit, k=3)

#Plotting dendogram with red borders around the 2 clusters
rect.hclust(fit, k=3, border="red")

gap_stat <- clusGap(scale, FUN = kmeans, nstart = 1, K.max = 6, B = 50)

fviz_gap_stat(gap_stat)

km.res3 <- kmeans(scale, 3, nstart = 10)  

# Visualize clusters
fviz_cluster(km.res3, data = scale,  
             ellipse.type = "convex",
             palette = "jco",
             ggtheme = theme_minimal())

```

# From the dendrogram, we can see that the countries can be grouped into three main clusters

```{r}
#•	Use principal components analysis to investigate the relationships between the countries based on these variables. Show visualizations to support your answers where possible

# Getting the Correlations 

animal_plot <- animalcor

corrplot(animal_plot,tl.cex = 0.7)

```

#Total score is least affected by (Kgs of pesticide/ Hectare of cropland) and affects most by Environmental Performance Index SCore and Support of Universal Declaration of ANimal Welfare
```{r}
animals.PCA <- prcomp(Animal_Rights[-1],scale=TRUE)
animals.PCA

str(animals.PCA)
#sdev,rotation,center,scale, x

animals.eigen <- animals.PCA$sdev^2
animals.eigen

sumlambdas <- sum(animals.eigen)
sumlambdas
#10

propvar <- animals.eigen/sumlambdas
propvar

cumvar <- cumsum(propvar)
cumvar

matlambdas <- rbind(animals.eigen,propvar,cumvar)
rownames(matlambdas) <- c("Eigenvalues","Prop. variance","Cum. prop. variance")
round(matlambdas,4)
summary(animals.PCA)
animals.PCA$rotation
print(animals.PCA)





plot(animals.eigen, xlab = "Component number", ylab = "Component variance", type = "l", main = "Scree diagram")

#cutoff for PCA dimension's should be at 2 the elbow break is massive after 2 dimension's indicating that the variance dropp off is massive

res.pca <- PCA(Animal_Rights[-1], graph = FALSE)

fviz_pca_var(res.pca, col.var = "black")

plot(log(animals.eigen), xlab = "Component number",ylab = "log(Component variance)", type="l",main = "Log(eigenvalue) diagram")

# Correlation
pairs.panels(Animal_Rights[,-1],
             gap = 0,
             bg = c("red", "blue")[Animal_Rights$Country],
             pch=21)

pairs.panels(animals.PCA$x,
             gap=0,
             bg = c("red", "blue")[Animal_Rights$Country],
             pch=21)

# Variables - PCA (cos2)
fviz_eig(animals.PCA, addlabels = TRUE)
fviz_pca_var(animals.PCA,col.var = "cos2",
             gradient.cols = c("#FFCC00", "#CC9933", "#660033", "#330033"),
             repel = TRUE)
# Individuals - PCA (cos2)
fviz_pca_ind(animals.PCA, col.ind = "cos2", 
             gradient.cols = c("#FFCC00", "#CC9933", "#660033", "#330033"), 
             repel = TRUE)




```

# Principal Component Analysis is used to reduce the large set of variables to a 
# small set that still contains most of the information from the large dataset and thereby, reducing the dimension of the dataset.
# After performing PCA on the ANimal_rights Dataset, we obtain 10 Principal components. 
# Each of these components represent the percentage of variability present in the dataset.
# In other words, PC1 explains 40.94% of total variance, PC2 explains 15.92% and so on. 
# We will consider the first two principal components as they sum up to 57% of total variance 
# and the others can be discarded as they contribute to only minimal amount of total variance.

```{r}
#	Identify the important factors underlying the observed variables and examine the relationships between the countries to these factors. Show visualizations to support your answers where possible.

# Note that here, we are excluding/ignoring the column 'total', because it is just having the summation of all the column values
fit.pc <- principal(Animal_Rights[c(-1,-11)], nfactors=6, rotate="varimax")
fit.pc

fit.pc <- principal(Animal_Rights[c(-1,-11)], nfactors=5, rotate="varimax")
fit.pc

fit.pc <- principal(animals[c(-1,-11)], nfactors=4, rotate="varimax")
fit.pc

# Factor recommendations for a simple structure
vss(Animal_Rights[c(-1,-11)]) 

# We are considering 4 factors, because of VSS (Very Simple Structure)
round(fit.pc$values, 4)



# Loadings
fit.pc$loadings
# Factor Loadings: This table displays the factor loadings for each variable. 
# It represents the correlation between each variable and each factor. 

# Communalities
fit.pc$communality
# Communalities are estimates of the variance in each observed variable that can be 
# explained by the extracted factors.
# It shows the proportion of each variable's variance that can be explained by the 


# Rotated factor scores
head(fit.pc$scores)
round(fit.pc$values,4)

# Factor recommendation
fa.parallel(Animal_Rights[c(-1,-11)])
# From this, we can inference that 4th component could be the best choice for number of factors

# Correlations within Factors
fa.plot(fit.pc)

# Visualizing the relationship
fa.diagram(fit.pc)


summary(vss(Animal_Rights[c(-1,-11)]) )

```
# Overall, the factor analysis suggests that there are four underlying factors that 
#The factor scores for each country provide a way to compare countries based on their animal rigts patterns and identify similarities and differences between them.
# Factor Analysis is the method of identifying the latent relational structure among 
# a set of variables and then narrowing it down to a smaller number of variables.
# We can see that we have reduced the factors to five, this contains most of the 
# information in the dataset.
# The factors RC1, RC3 can be helpful in analyzing the entire dataset.
#The parallel recommendations says two factors and two components should be taken.



#What would you name your principal components and factors? Why? 
.
#Factor 1: "Animal Rights and Environmental Advocacy"

#This factor represents attitudes and behaviors associated with advocating for animal rights and environmental conservation.
#It's characterized by variables with high loadings on both animal rights advocacy and environmental concerns, such as supporting animal welfare legislation and engaging in recycling initiatives.
#Factor 2: "Dietary Preferences and Consumption Habits"
#This factor captures attitudes and behaviors related to dietary preferences and consumption habits.
#It's characterized by variables with high loadings on dietary choices and consumption patterns, such as vegetarianism, organic food preferences, and frequency of meat consumption.
#For the Principal Component Analyses (PCAs):

#PCA 1: "Advocacy and Conservation Dimensions"

#This PCA explores the common dimensions underlying attitudes and behaviors related to advocacy for animal rights and environmental conservation.
#It aims to identify shared patterns of engagement in advocacy and conservation efforts.
#PCA 2: "Dietary and Consumption Patterns"

#This PCA focuses on variables related to dietary preferences and consumption habits.
#It aims to uncover underlying patterns in individuals' dietary choices and consumption behaviors, independent of advocacy and conservation concerns.

```{r}

```

